\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-07-15 tex/paper.tex
small edits

As a second toy example, we consider a generator producing
5-dimensional continuous data, as originally specified in Section 4.2. of
\citep{cranmer2015approximating}. More specifically, we consider the following
generative process:
\begin{itemize}
    \item $\bfz = (z_0, z_1, z_2, z_3, z_4)$, such that \\
        $z_0 \sim {\cal N}(\mu=\alpha, \sigma=1)$,
        $z_1 \sim {\cal N}(\mu=\beta, \sigma=3)$,
        $z_2 \sim {\text{Mixture}}[\frac{1}{2}\,{\cal N}(\mu=-2, \sigma=1), \\  \frac{1}{2}\,{\cal N}(\mu=2, \sigma=0.5)]$,
        $z_3 \sim {\text{Exponential}(\lambda=3)}$, and
        $z_4 \sim {\text{Exponential}(\lambda=0.5)}$;
    \item $\mathbf{x} = R  \bfz$, where $R$ is a fixed
    semi-positive definite $5 \times 5$ matrix defining a fixed projection
    of $\bfz$ into the observed space.
\end{itemize}
Again, the AVO algorithm does not have access to the density or its gradient, only samples from the generative model
We consider observed data generated at the nominal values $\bftheta^* = (\alpha^*=1,\beta^*=-1)$.
The simulator parameters are modeled with a factored  (mean field) Gaussian
variational distribution $q(\bftheta|\bfpsi) = q(\alpha|\bfpsi) q(\beta|\bfpsi)$, where each component was
initialized with zero mean and unit variance.
Hyper-parameters are set to $M=64$, $n_\text{critic}=100$, $\lambda=0.025$, $\gamma=10$ and
Adam configured with $\alpha=0.01$, $\beta_1=0.9$ and $\beta_2=0.999$.
The network architecture of the critic is the same as in the previous example.

@@

Adversarial Learned Inference (ALI)~\citep{dumoulin2016adversarially},
Bidirectional GANs (BiGANs)~\citep{donahue2016adversarial},
$\alpha$-GAN~\citep{rosca2017variational}, Adversarial Variational Bayes
(AVB)~\citep{DBLP:journals/corr/MeschederNG17}, and the \texttt{PC-Adv}
algorithm of  \citep{2017arXiv170208235H}  are a recent extension to GANs that
add an inference network to the generative model.  Each of these assume a
tractable density $p(\bfx|\bftheta)$ that is differentiable with respect to
$\bftheta$, which is  not satisfied in the likelihood-free setting. Our lifting
of the non-differentiable simulator $p(\bfx|\bftheta)$ to the variational
program $q(\bfx | \bfpsi)$ provides the ability to differentiate expectations
with respect to $\bfpsi$ as in Eqn~\ref{eqn:approx-grad}; however, the density
$q(\bfx | \bfpsi)$ is still intractable. Moreover, we do not attempt to define a
recognition model $q(\bfz | \bftheta, \bfx)$ as the latent space $\mathcal{Z}$
of many real-world simulators is complicated and not amenable to neural a
recognition model.
%One similarity of our work to AVB is that the ELBO is rewritten such that the intractable term $\log p(\bfz) - \log q_\bfphi(\bfz | \bfx)$ is replaced with an optimized neural network $T(x,z)$, which structurally similar to the role of the critic $d$ in the Kantorovich-Rubinstein dual description of the WGAN.

@@

From the point of view of likelihood-free inference, where  non-differentiable
simulators are the norm, our contributions are threefold. First is the process
of lifting expectations with respect to the non-differentiable simulator
$\mathbb{E}_{\tilde{\bfx} \sim p(\bfx | \bftheta)}$ to a differentiable
expectations with respect to the variational program $\mathbb{E}_{\tilde{\bfx}
\sim q(\bfx | \bfpsi)}$. Secondly, is the introduction of a novel form of
variational inference that works in a likelihood-free setting. Thirdly, the
AVI-MC algorithm can be viewed as a form of empirical Bayes where the prior is
optimized based on the data.

@@

Perhaps the closest to our work is~\cite{bernton2017inference}, which uses
Wasserstein distance both to find point estimates $\bftheta^*$ and as a part of
a rejection sampler in an ABC-like setup (as opposed to variational inference).
They emphasize that this approach allows one to eliminate the summary statistics
typically used in ABC and calculate the Wasserstein distance explicitly, without
making use of the Kantorovich-Rubinstein duality and of a critic $d$. For high-dimensional
data, they note that this is computationally expensive and introduce an
approximation based on projection of the data onto Hilbert space-filling curves.
Their Remark 5.1 points out that~\citep{montavon2016wasserstein} proposed an
approximation of the gradient of an entropy-regularized Wasserstein distance,
which uses a similar duality. They note that ``Unfortunately, it is not
applicable in the setting of purely generative models, as it involves point-wise
evaluations of the derivative of the log-likelihood.'' Thus, our contribution is
to provide gradients of an approximate MWE by taking expectations with the
variational program $q(\bfx | \bfpsi)$. This paired with the critic
Kantorovich-Rubinstein dual formulation of the Wasserstein distance
allows us to work in high dimensions without summary statistics and
optimize $\bfpsi$ with stochastic gradient descent. Additionally,
our our AVO-EB procedure avoids the inefficiencies of
their Wasserstein rejection sampler incurred from using the ABC-likelihood.

@@

This work has also many connections to work on variational inference, in which the
goal is to optimize the recognition model $q(\bfz, \bftheta|\bfpsi)$ so that it
is close to the true posterior $p(\bfz, \bftheta |\mathbf{x})$. There have been efforts
to extend variational inference to intractable likelihoods; however, they
require restrictive assumptions.  In \citep{tran2017variational}, the authors
consider Variational Bayes with an Intractable Likelihood (VBIL). In that
approach ``the only requirement is that the intractable likelihood can be
estimated unbiasedly.'' In the case of simulators, they propose to use the
ABC-likelihood with an $\epsilon$ kernel. The ABC likelihood is only unbiased as
$\epsilon > 0$,   thus this method inherits the drawbacks of the ABC-likelihood
including the choice of summary statistics and the inefficiency in evaluating
the ABC likelihood for high-dimensional data and small $\epsilon$.

@@

Lastly, we make a connection to Operator Variational Inference (OPVI)~\citep{2016arXiv161009033R},
which is a generalization of variational inference formulated as the following optimization problem
\begin{eqnarray}
\bflambda^* = \inf_\bflambda \sup_\bfphi \mathbb{E}_{ \bfz \sim q( \bfz | \bflambda)} [ (O^{p,q} f_\bfphi) ] \, .
\end{eqnarray}
In traditional VI with the KL distance, this corresponds to $(O^{p,q} f) = \log q(\bfz) - \log p(\bfz|\bfx) \, \forall f \in \mathcal{F}$.
AVO can be cast into a similar form with expectations over $\bfx$ instead of $\bfz$ and
\begin{eqnarray} \label{eq:opvi_avo}
\bfpsi^* &=& \inf_\bfpsi \sup_\bfphi \mathbb{E}_{ \bfx \sim q( \bfx | \bfpsi)} [ (O^{p_r,p_\bfpsi} d_\bfphi) ] \nonumber \\
\label{eq:opvi_avo2} &=&  \inf_\bfpsi \sup_\bfphi\mathbb{E}_{\bfx\sim p_r(\bfx)}[d(\bfx)] - \mathbb{E}_{\bfx \sim q(\bfx|\bfpsi)}[d(\bfx)] \, .
\end{eqnarray}
Rewriting Eqn.~\ref{eq:opvi_avo} as Eqn.~\ref{eq:opvi_avo2} is possible through importance sampling,
corresponding to an implicit form of the operator
\begin{equation}
(O^{p,q} d) = \left( \frac{p_r(\bfx)}{q(\bfx|\bfpsi)}  -1 \right) d(\bfx) \, ,
\end{equation}
which reinforces the link to density ratio estimation and inference with implicit models.

@@

In this note, we develop likelihood-free inference algorithms for non-differentiable, implicit generative models.
The algorithms combine adversarial training  with variational optimization to
minimize variational upper bounds  on the otherwise non-differentiable
adversarial objectives.
The AVO-EB algorithm enables empirical Bayesian through variational inference in the likelihood free setting.
This approach does not incur the inefficiencies of an ABC-like rejection sampler. The AVO-MWE algorithm
provides point estimates for the generative model, which asymptotically correspond to the data generating parameter when the model is well-specified. Both algorithms work on continuous or discrete observed data.

@@

\section*{Acknowledgments}

We would like to thank Lukas Heinrich for helpful comments regarding
the electron--positron annihilation simulation.

GL and KL are both supported through NSF ACI-1450310, additionally KC is
supported through PHY-1505463 and PHY-1205376.

>>>>

As a second toy example, we consider a generator producing
5-dimensional continuous data, as originally specified in Section 4.2. of
\citep{cranmer2015approximating}. More specifically, we consider the following
generative process:
\begin{itemize}
    \item $\bfz = (z_0, z_1, z_2, z_3, z_4)$, such that \\
        $z_0 \sim {\cal N}(\mu=\alpha, \sigma=1)$,
        $z_1 \sim {\cal N}(\mu=\beta, \sigma=3)$,
        $z_2 \sim {\text{Mixture}}[\frac{1}{2}\,{\cal N}(\mu=-2, \sigma=1), \\  \frac{1}{2}\,{\cal N}(\mu=2, \sigma=0.5)]$,
        $z_3 \sim {\text{Exponential}(\lambda=3)}$, and
        $z_4 \sim {\text{Exponential}(\lambda=0.5)}$;
    \item $\mathbf{x} = R  \bfz$, where $R$ is a fixed
    semi-positive definite $5 \times 5$ matrix defining a fixed projection
    of $\bfz$ into the observed space.
\end{itemize}
Again, the AVO algorithm does not have access to the density or its gradient, only samples from the generative model.
We consider observed data generated at the nominal values $\bftheta^* = (\alpha^*=1,\beta^*=-1)$.
The simulator parameters are modeled with a factored  (mean field) Gaussian
variational distribution $q(\bftheta|\bfpsi) = q(\alpha|\bfpsi) q(\beta|\bfpsi)$, where each component was
initialized with zero mean and unit variance.
Hyper-parameters are set to $M=64$, $n_\text{critic}=100$, $\lambda=0.025$, $\gamma=10$ and
Adam configured with $\alpha=0.01$, $\beta_1=0.9$ and $\beta_2=0.999$.
The architecture of the critic is the same as in the previous example.

@@

Adversarial Learned Inference (ALI)~\citep{dumoulin2016adversarially},
Bidirectional GANs (BiGANs)~\citep{donahue2016adversarial},
$\alpha$-GAN~\citep{rosca2017variational}, Adversarial Variational Bayes
(AVB)~\citep{DBLP:journals/corr/MeschederNG17}, and the \texttt{PC-Adv}
algorithm of  \citep{2017arXiv170208235H}  are a recent extension to GANs that
add an inference network to the generative model.  Each of these assume a
tractable density $p(\bfx|\bftheta)$ that is differentiable with respect to
$\bftheta$, which is  not satisfied in the likelihood-free setting. Our lifting
of the non-differentiable simulator $p(\bfx|\bftheta)$ to the variational
program $q(\bfx | \bfpsi)$ provides the ability to differentiate expectations
with respect to $\bfpsi$ as in Eqn~\ref{eqn:approx-grad}; however, the density
$q(\bfx | \bfpsi)$ is still intractable. Moreover, we do not attempt to define a
recognition model  $q(\bfz, \bftheta|\bfpsi)$ as the latent space $\mathcal{Z}$
of many real-world simulators is complicated and not amenable to neural a
recognition model.
%One similarity of our work to AVB is that the ELBO is rewritten such that the intractable term $\log p(\bfz) - \log q_\bfphi(\bfz | \bfx)$ is replaced with an optimized neural network $T(x,z)$, which structurally similar to the role of the critic $d$ in the Kantorovich-Rubinstein dual description of the WGAN.

@@

From the point of view of likelihood-free inference, where  non-differentiable
simulators are the norm, our contributions are threefold. First is the process
of lifting the expectation with respect to the non-differentiable simulator
$\mathbb{E}_{\tilde{\bfx} \sim p(\bfx | \bftheta)}$ to a differentiable
expectation with respect to the variational program $\mathbb{E}_{\tilde{\bfx}
\sim q(\bfx | \bfpsi)}$. Secondly, is the introduction of a novel form of
variational inference that works in a likelihood-free setting. Thirdly, the
AVO-EB algorithm can be viewed as a form of empirical Bayes where the prior is
optimized based on the data.

@@

Perhaps the closest to our work is~\cite{bernton2017inference}, which uses
Wasserstein distance both to find point estimates $\bftheta^*$ and as a part of
a rejection sampler in an ABC-like setup (as opposed to variational inference).
They emphasize that this approach allows one to eliminate the summary statistics
typically used in ABC and calculate the Wasserstein distance explicitly, without
making use of the Kantorovich-Rubinstein duality and of a critic $d$. For high-dimensional
data, they note that this is computationally expensive and introduce an
approximation based on projection of the data onto Hilbert space-filling curves.
Their Remark 5.1 points out that~\citep{montavon2016wasserstein} proposed an
approximation of the gradient of an entropy-regularized Wasserstein distance,
which uses a similar duality. They note that ``unfortunately, it is not
applicable in the setting of purely generative models, as it involves point-wise
evaluations of the derivative of the log-likelihood.'' Thus, our contribution is
to provide gradients of an approximate MWE by taking expectations with the
variational program $q(\bfx | \bfpsi)$. This paired with the critic
Kantorovich-Rubinstein dual formulation of the Wasserstein distance
allows us to work in high dimensions without summary statistics and
optimize $\bfpsi$ with stochastic gradient descent. Additionally,
our AVO-EB procedure avoids the inefficiencies of
their Wasserstein rejection sampler incurred from using the ABC-likelihood.

@@

This work has also many connections to work on variational inference, in which the
goal is to optimize the recognition model $q(\bfz, \bftheta|\bfpsi)$ so that it
is close to the true posterior $p(\bfz, \bftheta |\mathbf{x})$. There have been efforts
to extend variational inference to intractable likelihoods; however, they
require restrictive assumptions.  In \citep{tran2017variational}, the authors
consider Variational Bayes with an Intractable Likelihood (VBIL). In that
approach ``the only requirement is that the intractable likelihood can be
estimated unbiasedly.'' In the case of simulators, they propose to use the
ABC-likelihood with an $\epsilon$ kernel. The ABC likelihood is only unbiased as
$\epsilon \to 0$,   thus this method inherits the drawbacks of the ABC-likelihood
including the choice of summary statistics and the inefficiency in evaluating
the ABC likelihood for high-dimensional data and small $\epsilon$.

@@

Lastly, we make a connection to Operator Variational Inference (OPVI)~\citep{2016arXiv161009033R},
which is a generalization of variational inference formulated as the following optimization problem:
\begin{eqnarray}
\bflambda^* = \inf_\bflambda \sup_\bfphi \mathbb{E}_{ \bfz \sim q( \bfz | \bflambda)} [ (O^{p,q} f_\bfphi) ] \, .
\end{eqnarray}
In traditional VI with the KL distance, we have $(O^{p,q} f) = \log q(\bfz) - \log p(\bfz|\bfx) \, \forall f \in \mathcal{F}$.
AVO can be cast into a similar form with expectations over $\bfx$ instead of $\bfz$ and
\begin{eqnarray} \label{eq:opvi_avo}
\bfpsi^* &=& \inf_\bfpsi \sup_\bfphi \mathbb{E}_{ \bfx \sim q( \bfx | \bfpsi)} [ (O^{p_r,q_\bfpsi} d_\bfphi) ] \\
\label{eq:opvi_avo2} &=&  \inf_\bfpsi \sup_\bfphi \mathbb{E}_{\bfx\sim p_r(\bfx)}[d(\bfx; \bfphi)] - \mathbb{E}_{\bfx \sim q(\bfx|\bfpsi)}[d(\bfx; \bfphi)] \nonumber \, .
\end{eqnarray}
Rewriting Eqn.~\ref{eq:opvi_avo} as above is possible through importance sampling,
corresponding to an implicit form of the operator
\begin{equation}
(O^{p_r,q_\bfpsi} d_\bfphi)(\bfx) = \left( \frac{p_r(\bfx)}{q(\bfx|\bfpsi)}  -1 \right) d(\bfx; \bfphi) \, ,
\end{equation}
which reinforces the link to density ratio estimation and inference with implicit models.

@@

In this note, we develop likelihood-free inference algorithms for non-differentiable, implicit generative models.
The algorithms combine adversarial training  with variational optimization to
minimize variational upper bounds  on the otherwise non-differentiable
adversarial objectives.
The AVO-EB algorithm enables empirical Bayes through variational inference in the likelihood-free setting.
This approach does not incur the inefficiencies of an ABC-like rejection sampler. The AVO-MWE algorithm
provides point estimates for the generative model, which asymptotically correspond to the data generating parameter when the model is well-specified. Both algorithms work on continuous or discrete observed data.

@@

\vspace{0.25cm}
\subsection*{Acknowledgments}

We would like to thank Lukas Heinrich for helpful comments regarding
the electron--positron annihilation simulation.
GL and KL are both supported through NSF ACI-1450310, additionally KC is
supported through PHY-1505463 and PHY-1205376.
